\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-13 coqimpl.tex
Move a lot from sec3.2 to sec3.1

These are the core extensions we proposed to support \cref{fig:STLC-example}, and we adapt them into \cref{fig:STLC-example2}. We use "Family" and "EndFamily" to scope. "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.} Both "Field" and "FieldClaim" are only allowed to refer to the former fields, to avoid possible inconsistency mentioned in \ref{chg:consistency}.
Nested families are also allowed.

To resolve \ref{chg:extensible-inductive-type}, our first step is to \textit{decouple} the definition of recursor handlers and the checking of exhaustiveness. The users will define recursive handlers for one inductive type by aggregating the (function) handlers into one family. Then
"FRecursor" will be used to construct a recursor by refering to the handler family.
\mintinline{Coq}{FRecursor recur about T motive f } "using"
"handlers" will generate a new field "recur" that is really doing computation
(using elimination principle of the inductive type "T") with motive "f". The handlers
are located in the family "handlers", similar to a large pattern-matching statement. This way, the reuse and the extension of the case handlers will be delegated by the inheritance of the family. For example, in \cref{fig:STLC-example2},  "FTheorem" has a similar functionality but is expected to be used for theorem proving---the handler and the family will be automatically constructed during proving.

We can also use "Family" and "SealFamily" to scope a (nested) family. The resulting family only support direct inheritance or "Override" (i.e. we can only override a sealed family as a whole together)---we will later explain this design decision and its usage, especially on resolving \ref{chg:definition-relevant-reasoning}. What's more, "Override" family can only apply to sealed family as well (not allowed on normal family).

We can inherit a top-level family "parent" using \mintinline{Coq}{Family ... extends parent} as a new top-level family. As expected, "Override"
will override a field from the parent. "Inherit" can be used to inherit
nested families, inductive types, and fields. "Final" is used to claim that the decorated field is non-overridable, and later we will show how it is used to resolve \ref{chg:definition-relevant-reasoning}. The fields inside a family
are order-sensitive because we have dependent types here. So "Inherit
Until" is used if the user wants to inherit a range of fields.  "Extend Family" is used only inside a inheriting family. Thus nested
family can only be inherited or extended together with its belonging family
(i.e. ultimately top level family), thus our support of family polymorphism is
incomplete.

@@

Once the user closes a family with "EndFamily", they can access its fields by using module syntax. For example, they can access "A.a" to get "1" after "Family A" is defined.\YZ{1. move some of sec 3.2 ~> 3.1; 2. hint vaguely on how to do type checking in sec 3.1 (user interface).. metatheory part should also mention this part in sec 3.1} This is achieved by a complete compilation from a family to a module, via aggregating the list of the modules in the internal representation: the plugin will repeatedly using Coq's "Include" command as the "Module A" in \cref{fig:plugin-example1} illustrates. Now, the user can use "Module A" anywhere as they are using "Family A".
\YZ{The reader wonders why this aggregation is needed. 
Seems like the place to talk about how the programmer can use A outside Family A.}\EDJreply{Please check.}

@@

% explaining inheritance
\textbf{Implementing Inheritance.}
To resolve \ref{chg:consistency}, we make sure when inheritance happens, not only the content will be inherited, but the context of the corresponding field is also ``inherited''. Thus the mentioned informal example in \ref{chg:consistency} will not apply here because the overriding "f" is also in the context where "self" is empty thus "self.g" doesn't type-check.

To simplify the implementation, we can consider all families as
inheritance and the standalone families are extending the empty family.
Thus, we need to deal with only inheritance during interactive theorem proving.

There are three kinds of inheritance---extend, override, and inherit. Simiarly, we
use a list to encode inheritance data in the
implementation. The inheritance data will include the
information of (1) the ``type'' of the parent family, (2) a list of
operation indicating how each field of the parent will be dealt with
(either inherited or overriden), (3) the operations indicating newly
extended fields, and (4) the ``type'' of the children family.
The surface syntax for the three kinds of inheritance
are demonstrated in the exemplar \cref*{fig:plugin-example3}. The
implementation of overriding, inheritance, and extension is achieved by
simply manipulating the internal representation of a family
correspondingly---we can swap the module in the list for overriding,
retain the module in the list for inheritance, and add new module into
the list for extension. The compilation from a family (internal
representation) to a module still acts the same.

@@

\textbf{Reasoning Needs Non-Overridable.}
This compilation strategy
has already allowed fields to be type-dependent on former fields thus we can see concretely how
\ref{chg:definition-relevant-reasoning} appears in our current formulation and how "Final" can help to solve this challenge.

% However, the current implementation is not satisfactory enough. There are cases where we need to expose
% the concrete definition of a given field (especially when we are proving
% properties about it).

% For example, we may
% define $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field
% and intend to prove its commutativity in a following field $"comm"$.
% But to prove so, we need to know more than just a type interface of
% "add" but also its \textbf{concrete fixpoint definition}. Otherwise, in the
% future, once "add" is overridden into a subtraction function then
% "comm" will have problem to be inherited.\YZ{The sentence reads as if you want to enable overriding "add", whereas what you really want is to expose the definition equality about "add".}
% A more concrete example is in \cref{fig:plugin-example1}, consider the
% case we have "Field b : a = 1 := .." instead. Then our
% compilation and Coq's type checking will both fail: in the context of
% "b" (module type "a_6") we only know "a" is of type "nat" but nothing
% more (and this is necessary so that "a" can be overridden in the
% future).
Take the field "c" of "Family I" in \cref{fig:plugin-example3} as the
example. We can see that the first attempt of proving "a = 1" fails.
\textit{This should fail}---because we want to retain the possibility
to override "a" in the future, but the mere existence of the field "c"
will stop that, since "c" witnesses "a" fix to "1". \textit{This will fail} in our implementation---the proof of "c" will require a concrete definition of "a",
but our compilation will type check \mintinline{Coq}{c : a = 1 := eq_refl} in (a module with) the context where we only know \mintinline{Coq}{a : nat} thus failed.

To enable the exposure of the concrete definition,
our plugin provide "Final" keyword. Once we decorate it upon "a'", the second attempt \mintinline{Coq}{c : a' = 1 := eq_refl} will succeed. Contrary to "a", our "a'" is not overridable in "Family I2".

% To fix that, we need more information on "a" than its type---we
% need to expose the concrete definition "a = 1" to the context of "b".
%\YZ{I had a hard time working through these two paras. I guess they are
%saying that there is a need to expose definitional equalities, and
%therefore we need non-overridables. This causal flow should be made
%more explicit.}\EDJreply{Good suggestion. I try to add this casual flow into the first line of the paragraph. Please check and resolve if find satisfied.}

We implement "Final" by
exposing the whole definition directly into the compiled module type. Doing so we will prohibit overriding because
subtraction doesn't have the same definition as $add$. We require the
users to manual decorate field as `Final' so that the plugin will proceed with this special treatment.
%\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on talking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. Don't resolve it until I add it into Figure 2.}

\textbf{Special Fields: Overridable/Sealed Family Simulating Sigma Type.} 
Non-overridable fields are still not expressive enough to achieve the full power of dependent type programming. There are cases where overriding and
(definition-relevant) reasoning are both needed. For example, we may want a monotonic function as a overridable field member. We cannot split into two fields
\mintinline{Coq}{f : nat → nat} and \mintinline{Coq}{p : monotonic f},
because to prove
the latter we need to have concrete definition of "f" in the family type, which
makes "f" non-overridable at all. One option is to use sigma type
\mintinline{Coq}{{f | monotonic f}}, but that can be cumbersome when we have
a bundle of stuff with multiple refinements.

To resolve this, we provide \textit{family sealing} mechanism to simulate the sigma type. In a sealed family, every field will expose its definition inside the sealed family but a family as a whole will be sealed into a ``relaxed family type''. Thus we can override one sealed family as a whole and change the internal implementation as long as they can satisfy the ``relaxed family type''.

And the compiled module type of
"MonotonicF" will be fixed throughout the extension in different
families. Any overriding family will always have these two fields "f, p".\YZ{Can an overriding, sealed family define new fields not in its parent?}\EDJreply{I don't find this a problem but I didn't support it in my plugin. }

In our example, the seal family "I.MonotonicF" is actually simulating
interfaces as well.  We have
\mbox{``\mintinline{Coq}{Axiom non_impl : forall {T : Type}, T.}''},
and we assigned it to both "f" and "p", and later instantiate in
"I2" with concrete implementation by overriding this family
"I.MonotonicF".
% Basically, {a : int -> int; property : a is monotonic}, in this case
%   we may want to override "a" using different computation, but we still wnat
%   this refinement/property
The implementation of sealed family is not so different from that of the normal family, but we need to be careful about the compiled module type.

@@

We start with the programming interface. We use "FInductive" to define an extensible inductive type and in the children family we use "Extend Finductive ..." to extend the type with new constructors. In \cref{fig:plugin-example2}, we make the example of boolean and three-valued boolean.

>>>>

\textbf{Family and Defining Fields.} These are the core extensions we proposed to support \cref{fig:STLC-example}, and we adapt them into \cref{fig:STLC-example2}. We use "Family" and "EndFamily" to scope. "Field" is used to define new members in a family. "FieldClaim" is similar as "Field" but will invoke proof-mode in Coq and thus the user can use tactic programming to define new members.\footnote{We expect "FieldClaim" to be used in a \textit{proof-irrelevant} situation and thus whether overridable or not doesn't matter.} Both "Field" and "FieldClaim" are only allowed to refer to the former fields, to avoid possible inconsistency mentioned in \ref{chg:consistency}.
Nested families are also allowed.

The user can use "self__A.<path>" to access the former defined fields inside a surrounding defining "Family A".\footnote{Note that, in most of our examples in the paper, these "self__" prefix is too long and verbose so we omitted them.} Note that, "self__A" is only visible in the Vernacular commands from our plugin.\footnote{(In other words, using "Definition" from Coq will not invoke the plugin to prepare "self__A" to be accessible---most original Vernacular command from vanilla Coq will not do such preparation.)}  Once the user closes a \textbf{top} family with "EndFamily", they can access the fields of the closed family by using module syntax. For example, they can access "A.a" to get "1" after "Family A" is defined. This "A" is truly accessible anywhere by vanilla Coq without any plugin preparation, just like a mundane module.

Most programming experiences are preserved from Coq---each term inside the field definition is a well-typed Gallina term in vanilla Coq. Note that, a Family is not a first class value and thus we need to project out the Gallina terms from them.

% Inheritance here
\textbf{Inheritance on Fields.} "Family A extends B" has to appear at the top level (not inside any Family) and will start to construct the children "Family A" given the parent "Family B". Every field inside the inheriting family will be one of the three kinds of ``inheritance''---extending, overriding, and inheriting. The syntax of extending a new field is to directly use "Field", while overriding requires a prefix of "Override" before "Field f : .." and "Inherit f" will inherits field "f" from the parent.

To resolve \ref{chg:consistency}, we make sure when inheriting or overriding on "f" happens, the context of the field "f" is also ``inherited''. Thus the mentioned informal example in \ref{chg:consistency} will not apply here because the overriding "f" is also in the context where "self" is empty thus "self.g" doesn't type-check. Thus, before we inherit "g", the dependee "f" needs to be inherited.

% Talk about Final
\textbf{Final as non-overridable.} Every field is by default overridable, reminding us \ref{chg:definition-relevant-reasoning}.  To resolve it, we provide the keyword "Final" to allow the user to specify \textit{non-overridability} and give the user the opportunity to choose from overriding and exposing the definition.

Take the field "c" of "Family I" in \cref{fig:plugin-example3} as the
example. We can see that the first attempt of proving "a = 1" fails.
This should fail---because we want to retain the possibility
to override "a" and inheriting "c" in the future, but they are conflicting, since "c" witnesses "a" fix to "1". On the other hand, once we decorate it upon "a'", the second attempt \mintinline{Coq}{c : a' = 1 := eq_refl} will succeed. Contrary to "a", our "a'" is not overridable in "Family I2", mediating the \ref{chg:definition-relevant-reasoning}.

In our example, the seal family "I.MonotonicF" is actually simulating
interfaces as well.  We have
\mbox{``\mintinline{Coq}{Axiom non_impl : forall {T : Type}, T.}''},
and we assigned it to both "f" and "p", and later instantiate in
"I2" with concrete implementation by overriding this family
"I.MonotonicF".

\textbf{Overridable/Sealed Family as better Sigma Type.} Non-overridable fields are still not expressive enough to achieve the full power of dependent type programming. There are cases where overriding and
(definition-relevant) reasoning are both needed. For example, we may want a monotonic function as a overridable field member. We cannot split into two fields
\mintinline{Coq}{f : nat → nat} and \mintinline{Coq}{p : monotonic f},
because to prove
the latter we need to have concrete definition of "f" in the family type, which
makes "f" non-overridable at all. One option is to use sigma type
\mintinline{Coq}{{f | monotonic f}}, but that can be cumbersome when we have
a bundle of stuff with multiple refinements.

To resolve this, we provide \textit{family sealing} mechanism to simulate the sigma type. In a sealed family, every field will expose its definition inside the sealed family but a family as a whole will be ``sealed'' with a ``relaxed family type''---it is an aggregation of the types instead of the concrete definitions of all the field. Thus we can override one sealed family as a whole and change the internal implementation as long as they can satisfy the ``relaxed family type''.

% if the user wants a given field to be non-overridable. We will later elaborate its detail application.

\textbf{Inheritance on Nested Families.} We also provide extending, inheriting and overriding for nested family (member). Extension is done by simply declaring a nested family; the same syntax for inheriting a field is used in inheriting a family. However, overriding is special---overriding can only apply to \textit{sealed} nested family in the parent family. A family is sealed when using "SealFamily" as the scoping instead of "EndFamily". We will later elaborate the motivation and details of sealed family.

% Currently, every sealed family can only be direct inherited or overridden as a whole. It cannot be extended.

\textbf{Extensible Inductive Type.} We provide the keyword "FInductive" to enable extensible inductive type. "Extend FInductive" is used to support extension of the parent inductive type inside a children family. The syntax of these two is close to the mundane "Inductive" keyword.

To resolve \ref{chg:extensible-inductive-type}, our first step is to \textit{decouple} the definition of recursor handlers and the checking of exhaustiveness. The users will define recursive handlers for one inductive type by aggregating the (function) handlers into one family. Then
"FRecursor" will be used to construct a recursor by refering to the handler family.
\mintinline{Coq}{FRecursor recur about T motive f } "using"
"handlers" will generate a new field "recur" that is really doing computation
(using elimination principle of the inductive type "T") with motive "f". The handlers
are located in the family "handlers", similar to a large pattern-matching statement. This way, the reuse and the extension of the case handlers will be delegated by the inheritance of the family. Once recursor is created it will look like any other function fields(e.g. "subst" in \cref{fig:STLC-example2}).

"FTheorem" has a similar functionality but is expected to be used for theorem proving---the handler and the family will be automatically constructed during proving.

% We can also use "Family" and "SealFamily" to scope a (nested) family. The resulting family only support direct inheritance or "Override" (i.e. we can only override a sealed family as a whole together)---we will later explain this design decision and its usage, especially on resolving \ref{chg:definition-relevant-reasoning}. What's more, "Override" family can only apply to sealed family as well (not allowed on normal family).

% We can inherit a top-level family "parent" using \mintinline{Coq}{Family ... extends parent} as a new top-level family. As expected, "Override"
% will override a field from the parent. "Inherit" can be used to inherit
% nested families, inductive types, and fields. "Final" is used to claim that the decorated field is non-overridable, and later we will show how it is used to resolve \ref{chg:definition-relevant-reasoning}. The fields inside a family
% are order-sensitive because we have dependent types here. So "Inherit
% Until" is used if the user wants to inherit a range of fields.  "Extend Family" is used only inside a inheriting family. Thus nested
% family can only be inherited or extended together with its belonging family
% (i.e. ultimately top level family), thus our support of family polymorphism is
% incomplete.

@@

\YZ{1. move some of sec 3.2 ~> 3.1; 2. hint vaguely on how to do type checking in sec 3.1 (user interface).. metatheory part should also mention this part in sec 3.1} This is achieved by a complete compilation from a family to a module, via aggregating the list of the modules in the internal representation: the plugin will repeatedly using Coq's "Include" command as the "Module A" in \cref{fig:plugin-example1} illustrates. Now, the user can use "Module A" anywhere as they are using "Family A".
\YZ{The reader wonders why this aggregation is needed. 
Seems like the place to talk about how the programmer can use A outside Family A.}\EDJreply{Please check.}

@@

% explaining inheritance
\textbf{Implementing Inheritance.}
To simplify the implementation, we can consider all families as
inheritance and the standalone families are extending the empty family.
Thus, we need to deal with only inheritance during interactive theorem proving.

Simiarly, we
use a list to encode the three kinds of inheritance data in the
implementation. The inheritance data will include the
information of (1) the ``type'' of the parent family, (2) a list of
operation indicating how each field of the parent will be dealt with
(either inherited or overriden), (3) the operations indicating newly
extended fields, and (4) the ``type'' of the children family.
The surface syntax for the three kinds of inheritance
are demonstrated in the exemplar \cref*{fig:plugin-example3}. The
implementation of overriding, inheritance, and extension is achieved by
simply manipulating the internal representation of a family
correspondingly---we can swap the module in the list for overriding,
retain the module in the list for inheritance, and add new module into
the list for extension. The compilation from a family (internal
representation) to a module still acts the same.

@@

Our implementation also make sure the first attempt of defining "c" failed successfully in \cref{fig:plugin-example3}---the proof of "c" will require a concrete definition of "a",
but our compilation will type check \mintinline{Coq}{c : a = 1 := eq_refl} in (a module with) the context where we only know \mintinline{Coq}{a : nat} thus failed.

\textbf{Implemening "Final" and "Sealed Family".}
We implement "Final" by
exposing the whole definition directly into the compiled module type. Doing so will prohibit overriding. Say, the example "bop" in \cref{fig:plugin-example3}, will be prohibited to be overridden by $(-)$
subtraction doesn't have the same definition as $(+)$. We require the
users to manual decorate field as `Final' so that the plugin will proceed with this special treatment.
%\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on talking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. Don't resolve it until I add it into Figure 2.}

% \textbf{Special Fields: Overridable/Sealed Family Simulating Sigma Type.}

The implementation of sealed family is not so different from that of the normal family, but we need to deal with the compiled module type. The compiled module type is the aggregation of the type instead of the concrete definition of the fields.
For example, the compiled module type of
"MonotonicF" in \cref{fig:plugin-example3} will be "{f : nat -> nat; p : monotonic f}". Thus any future overriding family will always have these two fields "f, p" with the corresponding type.\YZ{Can an overriding, sealed family define new fields not in its parent?}\EDJreply{I don't find this a problem but I didn't support it in my plugin. }

% Basically, {a : int -> int; property : a is monotonic}, in this case
%   we may want to override "a" using different computation, but we still wnat
%   this refinement/property

@@

% We start with the programming interface. We use "FInductive" to define an extensible inductive type and in the children family we use "Extend Finductive ..." to extend the type with new constructors. In \cref{fig:plugin-example2}, we make the example of boolean and three-valued boolean.
